NanoMaLy

### 2015-11-19 paper/pldi16-submission/semantics.tex
down to 11pp

\paragraph{Narrowing Types} The procedure $\force{e}{t}{\vsu}$ takes as
input an term \texttt{e} and refines it to have the type \texttt{t} by
yielding a pair of either the same expression and a value substitution
the refines the holes in \texttt{e} to ensure it has type \texttt{t}, or
yields the stuck term if no such refinement is possible. In the case
where \texttt{e} is a hole, it first checks in the given $\vsu$ to see
if the hole has already been instantiated and, if so, returns the
existing instantiation:
%
$$
\begin{array}{lcl}
% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
% \forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
\force{\ehole{i}}{t}{\vsu} & \defeq & \hspace{-1ex} \begin{array}{l}
    \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \pair{\vsu(i)}{\emptysu}\ \mathtt{else} \\
         \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}}
  \end{array} \\
\force{v}{\ehole{}}{\vsu}  & \defeq & \pair{v}{\emptysu} \\
\force{n}{\tint}{\vsu}     & \defeq & \pair{n}{\emptysu} \\
\force{v}{\tint}{\vsu}     & \defeq & \pair{\stuck}{\emptysu} \\
\force{b}{\tbool}{\vsu}    & \defeq & \pair{b}{\emptysu} \\
\force{v}{\tbool}{\vsu}    & \defeq & \pair{\stuck}{\emptysu} \\
\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\efun{x}{e}}{\emptysu} \\
\force{v}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\stuck}{\emptysu} \\
\end{array}
$$

@@

\paragraph{Preservation}
Given these two definitions, we show that each evaluation step
refines the partial type of $f$, \ie preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \pair{\eapp{f}{\ehole{}}}{\emptysu},\ldots$ and
$\trace' \defeq \trace, \pair{e'}{\vsu'}$ (\ie $\trace'$ is single
step extension of $\trace$)
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
\end{lem}
\begin{proof}
  By induction on the length of $\trace$ and case analysis on the evaluation rules.
  %
  Note that all rules preserve partial types with the exception of when
  \forcesym is called on a hole, in which case we instantiate the hole with
  a concrete value.
  %
  But \hastype{\ehole{}}{\thole{}}, which is compatible with any type.
\end{proof}

@@

\paragraph{Incompatible Values Are Wrong}
%
We show that \emph{any} value that is \emph{incompatible} with
the partial type upto $\trace$ will cause $f$ to get stuck
in \emph{at most} $k+1$ steps where $k$ is the length of the
trace $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  If \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\vsu},
     \tincompat{t}{\ptype{\trace}{f}}, and
     \hastype{v}{t},
  then
     \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
\end{lem}
\ES{self-todo: take a close look at the $\trace$'s vs k's}
\begin{proof}
  By induction on $k$, the length of $\trace$.

  Let $\tfun{s_{\trace}}{\_} = \ptype{\trace}{f}$.
  Suppose \hastype{v}{\tincompat{t}{s_{\trace'}}}, we
  will show that \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su'}
  in at most $k+1$ steps:
  \begin{description}
  \item[Case \tincompat{t}{s_{\trace}}:]
    The inductive hypothesis applies.
  \item[Case \tcompat{t}{s_{\trace}} but \tincompat{t}{s_{\trace'}}:]
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k+1$.
    %
    A case analysis of the applicable rules shows that \forcesym cannot
    have succeeded.
  \end{description}
\end{proof}

\paragraph{Proof of Theorem~\ref{thm:soundness}}
%
Now we can prove our soundness theorem. Intuitively, suppose $\trace$
witnesses that $f$ gets stuck, and let $t$ be the partial type of $f$
upto $\trace$. We show that \emph{all} types $s$ have stuck-inducing
values by splitting cases on whether the type is compatible with the
partial type upto $\trace$.
%
\begin{description}
  \item[Case \tcompat{s}{t}:]
   The value $\hastype{v}{t}$ suffices to demonstrate that $\eapp{f}{v}$
   gets stuck.
  \item[Case \tincompat{s}{t}:]
   By Lemma~\ref{lem:k-stuck}, every \hastype{v}{t} demonstrates that
   $\eapp{f}{v}$ gets stuck.
\end{description}

@@

\begin{figure}[ht]
\centering
\begin{mcode}
genArgs :: [($x$, $e$)] -> [$v$]
genArgs bs = loop []
  where
  loop vs  = case eval (close bs vs) of
               $\pair{\efun{x}{e}}{\_}$:_ -> loop ($\ehole{}$:vs)
               _        -> vs

@@

\begin{figure}[ht]
\centering
\begin{mcode}
genWitness :: Nat -> [($x$, $e$)] -> [$v$]
genWitness n bs
       = [witness bs ($\vsu$ vs) | $\vsu$ <- $\vsu$s]
  where
   vs  = genArgs defs []      -- (1)
   $e$    = close defs vs        -- (2)
   res = take n (eval e)      -- (3)
   $\vsu$s   = [$\vsu$ | $\pair{\stuck}{\vsu}$ <- res]  -- (4)
\end{mcode}
\caption{Generating Witnesses}
\label{fig:algo-gen-witness}
\end{figure}

>>>>

\paragraph{Narrowing Types} The procedure $\force{e}{t}{\vsu}$ takes as
input an term \texttt{e} and refines it to have the type \texttt{t} by
yielding a pair of either the same expression and a value substitution
the refines the holes in \texttt{e} to ensure it has type \texttt{t}, or
yields the stuck term if no such refinement is possible. In the case
where \texttt{e} is a hole, it first checks in the given $\vsu$ to see
if the hole has already been instantiated and, if so, returns the
existing instantiation:
%
$$
\begin{array}{lcl}
% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
% \forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
\force{\ehole{i}}{t}{\vsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \pair{\vsu(i)}{\emptysu}        & \mbox{if}\ i \in \vsu \\
  \pair{v}{\ehole{i} \mapsto v}   & \mbox{o.w.} v = \gen{t}\\
\end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \pair{\vsu(i)}{\emptysu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{v}{\ehole{}}{\vsu}  & \defeq & \pair{v}{\emptysu} \\
\force{n}{\tint}{\vsu}     & \defeq & \pair{n}{\emptysu} \\
\force{v}{\tint}{\vsu}     & \defeq & \pair{\stuck}{\emptysu} \\
\force{b}{\tbool}{\vsu}    & \defeq & \pair{b}{\emptysu} \\
\force{v}{\tbool}{\vsu}    & \defeq & \pair{\stuck}{\emptysu} \\
\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\efun{x}{e}}{\emptysu} \\
\force{v}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\stuck}{\emptysu} \\
\end{array}
$$

@@

\paragraph{Preservation}
Given these two definitions, we show that each evaluation step
refines the partial type of $f$, \ie preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \pair{\eapp{f}{\ehole{}}}{\emptysu},\ldots$ and
$\trace' \defeq \trace, \pair{e'}{\vsu'}$ (\ie $\trace'$ is single
step extension of $\trace$)
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
\end{lem}
\begin{proof}
  By induction on the length of $\trace$ and case analysis on the evaluation rules.
  %
  Note that all rules preserve partial types with the exception of when
  \forcesym\ is called on a hole, in which case we instantiate the hole with
  a concrete value.
  %
  But \hastype{\ehole{}}{\thole{}}, which is compatible with any type.
\end{proof}

@@

\paragraph{Incompatible Values Are Wrong}
%
\emph{Any} value that is \emph{incompatible} with
the partial type upto trace $\trace$ will cause $f$ to get stuck
in \emph{at most} $k+1$ steps where $k$ is the length of $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  If \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\vsu},
     \tincompat{t}{\ptype{\trace}{f}}, and
     \hastype{v}{t},
  then
     \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
\end{lem}
\ES{self-todo: take a close look at the $\trace$'s vs k's}
\begin{proof}
  By induction on $k$, the length of $\trace$.
  Let $\tfun{s_{\trace}}{\_} = \ptype{\trace}{f}$.
  Suppose \hastype{v}{\tincompat{t}{s_{\trace'}}}, we
  will show that \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su'}
  in at most $k+1$ steps:
%
  %\begin{description}
  % \item
  [Case \tincompat{t}{s_{\trace}}:]
    The inductive hypothesis applies.
  %\item
  [Case \tcompat{t}{s_{\trace}} but \tincompat{t}{s_{\trace'}}:]
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k+1$.
    %
    A case analysis of the applicable rules shows that \forcesym cannot
    have succeeded.
  %\end{description}
\end{proof}

\paragraph{Proof of Theorem~\ref{thm:soundness}}
%
Suppose $\trace$ witnesses that $f$ gets stuck,
and let $t$ be the partial type of $f$ upto $\trace$.
We show that \emph{all} types $s$ have stuck-inducing
values by splitting cases on whether the type is
compatible with the partial type upto $\trace$.
%
% \begin{description}
  %\item
  [Case \tcompat{s}{t}:]
   The value $\hastype{v}{t}$ suffices to demonstrate that $\eapp{f}{v}$
   gets stuck.
  %\item
  [Case \tincompat{s}{t}:]
   By Lemma~\ref{lem:k-stuck}, every \hastype{v}{t} demonstrates that
   $\eapp{f}{v}$ gets stuck.
%\end{description}

@@

\begin{figure}[t]
\centering
\begin{mcode}
genArgs :: [($x$, $e$)] -> [$v$]
genArgs bs = loop []
  where
  loop vs  = case eval (close bs vs) of
               $\pair{\efun{x}{e}}{\_}$:_ -> loop ($\ehole{}$:vs)
               _        -> vs

@@

\begin{figure}[t]
\centering
\begin{mcode}
genWitness :: Nat -> [($x$, $e$)] -> [$v$]
genWitness n bs
       = [witness bs ($\vsu$ vs) | $\vsu$ <- $\vsu$s]
  where
   vs  = genArgs defs []      -- (1)
   $e$    = close defs vs        -- (2)
   res = take n (eval e)      -- (3)
   $\vsu$s   = [$\vsu$ | $\pair{\stuck}{\vsu}$ <- res]  -- (4)
\end{mcode}
\caption{Generating Witnesses}
\label{fig:algo-gen-witness}
\end{figure}
